```{r setup-Logistische-Regression-Englisch, include=FALSE}
source("../prelude.R")
# ---------------------------------------------------------------------------
#% maintainer:
#%   - Karsten Luebke
#%
# ---------------------------------------------------------------------------
source("../prelude.R")
initPart(
    "LogistischeRegression",  # Dateiname ohne Suffix
    "Regression"                # Verzeichnisname im Bilderverzeichnis 
    )
pathToImages = getPathToImages()
# ---------------------------------------------------------------------------

library(mosaic)
library(knitr)

tips <- assertData("tips.csv", "https://goo.gl/whKjnl")
```
# Logistic Regression




### Exercise `r nextExercise()`: Scale level `smoker`  {.exercise type=A-B-C-D answer=A}

What's the scale level of `smoker` (values `Yes` and `No`)?


A.  Categorical - nominal.
B.  Categorical - ordinal.
C.  Numeric - discrete.
D.  Numeric - continuous.


::: {.notes}
***A***
:::



### Statistical modelling and logistic regression

$$y=f(x) + \epsilon $$

Let's start with one independent variable only:

- Linear Regression: Dependent Variable $y$ numeric: $y_i=\beta_0 + \beta_1 \cdot x_i + \epsilon_i$
- **Logistische Regression**: **Dependent Variable** $y$ **binary**, ie., categorical with two levels $y_i \in \{0,1\}$. 

Let $p_i$ the probability that $y_i=1$, then:

$logit(p_i)=\ln(\frac{p_i}{1-p_i})=\beta_0 + \beta_1 \cdot x_i$.  


- In R: `glm(y ~ x, data = my_data,  family='binomial')`


### Examples

- Will a given student pass the next exam successfully?
- Will a given coworker quit within the next six months?
- Will a given product being bought, given some customers data?
- Will a case become a "high cost case" given some data on the case and on the customer?

*For what question could you employ this method?*

### Setup 

Load the `tips` data set^[Bryant, P. G. and Smith, M (1995) Practical Data Analysis: Case Studies in Business Statistics. Homewood, IL: Richard D. Irwin Publishing]

Start the package `mosaic`.


```{r, eval= FALSE, message=FALSE}
download.file("https://goo.gl/whKjnl", destfile = "tips.csv")
tips <- read.csv2("tips.csv")

library(mosaic) # load the package 
```


### What's the probability that a man pays the bill?


... conditional on the amount of the bill (`total_bill`)?



```{r, fig.align="center", out.width="40%"}
gf_point( (sex=="Male") ~ total_bill, 
        data = tips)
```

Note: The gender distribution at a table is unknown. If for example there are only women or only men at the table, then the probability will be $0$ ($1$).


### Introducing the Logit

$$p(y=1)=\frac{e^\eta}{1+e^\eta}=\frac{e^{\beta_0 + \beta_1 \cdot x}}{1+e^{\beta_0 + \beta_1 \cdot x}} = \frac{1}{1+e^{-(\beta_0 + \beta_1 \cdot x)}}$$
```{r, echo=FALSE, fig.align="center", out.width="33%"}
eta <- seq(-10,10,by = 0.1)
y <- 1/(1+exp(-eta))
gf_line(y ~ eta, xlab = expression(eta), ylab = "p(y=1)")  
```

Estimate $\beta$ based on the data: $\hat{\beta}$.


- $\beta_1>0$: Probability $p(y=1)$ increases with $x$.
- $\beta_1<0$: Probability $p(y=1)$ decreases with $x$.


### Preparing the data: Set reference factor level

In R' logistic regression, $y$ is modelled according to the *factor levels*. The first level is assigned to $0$, the other levels are assigned to $1$.^[NB: Use the $-operator to call a variable in a data frame: my_data_frame$my_variable.] 

```{r}
# define reference level: 
tips$sex<- relevel(tips$sex, ref = "Female")
# check:
levels(tips$sex)
```


### Logistic Regression: Sex to total bill (I/II)

```{r, eval=FALSE}
# save output of regression as "my_glm1"
my_glm1 <- glm(sex ~ #  dependent variable (dv) 
               total_bill, # independent variables (iv) 
              data = tips, # data set 
              # DV is binomial
              # Link function:  Logit
              family = binomial("logit"))
              

summary(my_glm1)
```




### Logistic Regression: Sex to total bill (II/II) {.shrink}


```{r, echo=FALSE}
# save output of regression as "my_glm1"
my_glm1 <- glm(sex ~ # abh채ngige Variable
                 total_bill, # unabgh채ngige Variable(n)
              data = tips, # Datensatz
              # Abh채ngige Variable binomial,
              # Verkn체pfung Logit
              family = binomial("logit"))
              

summary(my_glm1)
```





### Output: Logistics Regression sex to total bill

```{r, fig.align="center", out.width="66%"}
plotModel(my_glm1)
```


### Exercise `r nextExercise()`: regression coefficients  {.exercise type=A-B-C answer=A}

Which proposition is true?


A.  According to the model, the probability of a man paying increases with the total bill.
B.  According to the model, the probability of a man paying decreases with the total bill.

C.  According to the model, the probability of a man paying is not associated  with the total bill.



::: {.notes}
**A**
:::



### Modelling sex by number of persons  {.shrink}

```{r}
ergglm2 <- glm(sex ~ size, 
              data = tips,
              family = binomial("logit"))
              
summary(ergglm2)
```


### Exercise `r nextExercise()`: Results: sex by number of persons {.exercise type=A-B-C answer=A}

At what level of `size` is the probability maximal that a *female* pays the bill?

A.  $1$ individual.
B.  $2$ individuals.
C.  $4$ individuals.

::: {.notes}
As $\hat{\beta}_1=`r round(coef(ergglm2)[2],2)`>0$ holds, the probability that a *man* pays increases with total bill increasing. Thus, the probability that a women pays is maximized when there the least number of individuals at the table, hence ***A***.
:::


### Predicting and the logistics regression

What's the probability that a man pays, given that there are $4$ persons:


```{r}
predict(ergglm2, 
        newdata = data.frame(size = 4), 
        type = "response")
```


### Model assessment

Typical model assessment for classification models include:

- accuracy: proportion of correctly classified cases
- Cohen's Kappa: Comparison of observed accuracy to expected accuracy (expected under independency)


### Modelling sex by time {.shrink}

```{r}
ergglm3 <- glm(sex ~ time, data = tips,
              family = binomial("logit"))
summary(ergglm3)
```


### Exercise `r nextExercise()`: Results -- Modelling sex by time {.exercise type=A-B-C answer=B}

According to the model, when is the probability maximized that a man pays the bill?


A.  Lunch.
B.  Dinner.
C.  No difference.

::: {.notes}
 $\hat{\beta}_{timeLunch}=`r round(coef(ergglm3)[2],2)`<0$.Hence ***B***.
:::


### Odds Ratio^[Chance vs. probability: The *probability* $p$ to roll a 6 (assuming a fair die) is $\frac{1}{6}$. The *chance* $\frac{\textrm{p}}{\textrm{1-p}}$, assuming a fair die, amounts to  $\frac{1}{5}$.]





+------------------------+-----------------------+---------------------------+
|                        | **Variable B: yes **  | **Variable B: no **       |
+========================+=======================+===========================+
| **Variable A: yes **   | eg., $n_{11}=100$     |  eg., $n_{12}=200$        |
+------------------------+-----------------------+---------------------------+
| **Variable A: no **    | eg., $n_{21}=150$     |  eg., $n_{22}=550$        | 
+------------------------+-----------------------+---------------------------+


```{r echo = FALSE, message = FALSE, eval = TRUE, cache = FALSE}
d <- read_csv("../Inhalte/StatisticalLearning/class-results.csv")



d %>% 
  kable(escape = FALSE, booktabs = TRUE) %>% 
  kable_styling() %>% 
  column_spec(1, width = "1.8cm") %>% 
  column_spec(2, width = "1.8cm") %>% 
  column_spec(3, width = "4cm") %>% 
  column_spec(3, width = "4cm")
```





- The odds for "yes" are $\frac{n_{11}+n_{12}}{n_{21}+n_{22}}=\frac{100+200}{150+550}$.




- The odds ratio for "yes" at $A$, if $B$ is "yes" too, amounts to: 
$$OR=\frac{\frac{n_{11}}{n_{21}}}{\frac{n_{12}}{n_{22}}}=\frac{\frac{100}{150}}{\frac{200}{550}}=\frac{55000}{30000}\approx1.83$$


### Odds Ratio and the logistic regression

Odds ratio:

$$OR=\frac{\frac{p_{\text{Lunch}}}{1-p_{\text{Lunch}}}}{\frac{p_{\text{Dinner}}}{1-p_{\text{Dinner}}}}=e^{\beta_{\text{time}}}$$
```{r}
exp(coef(ergglm3))
```

The *chance* that a man pays at lunch is `r round(exp(coef(ergglm3))[2],2)` times larger than at dinner (ie., at lunch the chance is *lower* than at dinner that a man pays). 



### Exercise `r nextExercise()`: Odds Ratio {.exercise type=A-B-C answer=C}


What's the odds ratio indicating that the chances are the same in both situations?


A.   $OR=0$
B.   $OR=0.5$
C.   $OR=1$

::: {.notes}
***C*** 
:::


### Multiple logistic regression

```{r}
ergglm4 <- glm(sex ~ 
                 total_bill + size + time,
              data = tips,
              family = binomial("logit"))
              
exp(coef(ergglm4)) # Odds Ratio
```


### Coefficients: Multiple logistic regression {.shrink}

```{r}
summary(ergglm4)
```


### Exercise `r nextExercise()`: Results -- Multiple logistic regression {.exercise type=A-B-C-D-E answer=D}

Which of the following options correctly indicates that a man pays the bill?



A.  Positive coefficient for total bill only.
B.  Negative coefficient for number of persons.
C.  Dinner.
D.  Positive coefficient for total bill, negative coefficient for number of persons, and dinner.
E.  None of the abov None of the abovee 

::: {.notes}
***D***.
:::



### Exercise `r nextExercise()`: Modelling sex  {.exercise type=essay}


Come up with a model to predict the probability that a man pays the bill using the predictors `total_bill`, `smoker` and `day`.


```{r, include=FALSE}
ergglmUeb <- glm(sex ~ total_bill + smoker + day, data = tips, family = binomial("logit"))
summary(ergglmUeb)
exp(coef(ergglmUeb))
```

::: {.notes}
`ergglmUeb <- glm(sex ~ total_bill + smoker + day, data = tips, family = binomial("logit"))`  
`summary(ergglmUeb)`  
`exp(coef(ergglmUeb))` 

:::


```{r finish-Logistische-Regression, include=FALSE}
rm(pathToImages)
finalizePart(partname)
```
