```{r, include=FALSE}
# ---------------------------------------------------------------------------
#% maintainer:
#%   - Sebastian Sauer
#%
# ---------------------------------------------------------------------------
source("../prelude.R")
initPart(
    "DES-Organisatorisches",  # Dateiname ohne Suffix
    "Organisatorisches"              # Verzeichnisname im Bilderverzeichnis 
    )
pathToImages <- getPathToImages()
# ---------------------------------------------------------------------------

```

# `r if(abschluss) WDH` Organizational notes 

```{r child = './Organisatorisches/Literatur-Englisch.Rmd'}
```

```{r child = './Organisatorisches/LizenzVersion-Englisch.Rmd'}
```




```{r child = './Organisatorisches/Vorlesungsplan.Rmd', eval = showVorlesungsplan}
```


### Upfront preparation


- Please download all materials upfront.
- Please make sure that you bring all materials to class at each time.
- Please make sure that you have access to the internet at all times.
- Please bring your computer with you to each class.
- Please make sure that you install R and RStudio before the first class.
- Please make sure that you have installed all R packages indicated in the course materials.
- Please make sure that you run uptodate versions of your software (including OS and R).

In case of any questions, do not hesitate to contact your instructor.






```{r, include=FALSE}
abschluss <<- TRUE
rm(pathToImages)
```
